### 2024-07-12 actuarial.qmd
20240712 - actuarial

# Human Judgment Versus Actuarial Approaches to Prediction {#sec-judgmentVsActuarial}

@@

Using the judgment method of prediction, all gathered information is collected and formulated into a prediction in the person's mind.
The person selects, measures, and combines information and produces projections solely according to their experience and judgment.

@@

$$
\text{Violence risk} = \beta \cdot \text{conduct disorder} + \beta \cdot \text{substance use} + \beta \cdot \text{suspended from school} + \beta \cdot \text{childhood aggression} + ...
$$ {#eq-actuarial}

@@

An actuarial formula for projecting a Running Backs rushing yards might be something like @eq-actuarialFantasy:

$$
\text{rushing yards} = \beta \cdot \text{rushing yards last season} + \beta \cdot \text{age} + \beta \cdot \text{injury history} + \beta \cdot \text{strength of offensive line} + ...
$$ {#eq-actuarialFantasy}

@@

The challenge is what to do when the human and the algorithm disagree.
Hypothetically, humans reviewing and adjusting the results from the statistical algorithm could lead to more accurate prediction.
However, human input also could lead to the possibility or exacerbation of biased predictions.
In general, with very few exceptions, actuarial approaches are as accurate or more accurate than "expert" judgment [@Dawes1989; @Aegisdottir2006; @Baird2000; @DenHartigh2018; @Grove1996; @Grove2000].
Moreover, the superiority of actuarial approaches to human judgment tends to hold even when the expert is given more information than the actuarial approach [@Dawes1989].
Allowing experts to override actuarial predictions consistently leads to lower predictive accuracy [@Garb2019].

There is sometimes a misconception that formulas cannot account for qualitative information.
However, that is not true.
Qualitative information can be scored or coded to be quantified so that it can be included in statistical formulas.
That said, the quality of predictions rests on the quality and relevance of the assessment information for the particular prediction decision.
If the assessment data are lousy, it is unlikely that a statistical algorithm (or a human for that matter) will make an accurate prediction:
"Garbage in, garbage out".
A statistical formula cannot rescue inaccurate assessment data.

>>>>

# Judgment Versus Actuarial Approaches to Prediction {#sec-judgmentVsActuarial}

@@

Using the judgment method of prediction, all gathered information is collected and formulated into a prediction in the person's mind.
The person selects, measures, and combines information and produces projections solely according to their experience and judgment.
For instance, a proclaimed "fantasy expert" might use their experience, expertise, and judgment to make a prediction about how each player will perform by using whatever information and data they deem to be important, aggregating all of this information in their mind to make the prediction for each player.

@@

$$
\scriptsize
\text{Violence risk} = \beta \cdot \text{conduct disorder} + \beta \cdot \text{substance use} + \beta \cdot \text{suspended from school} + \beta \cdot \text{childhood aggression} + ...
$$ {#eq-actuarial}

@@

An actuarial formula for projecting a Running Back's rushing yards might be something like @eq-actuarialFantasy:

$$
\scriptsize
\text{rushing yards} = \beta \cdot \text{rushing yards last season} + \beta \cdot \text{age} + \beta \cdot \text{injury history} + \beta \cdot \text{strength of offensive line} + ...
$$ {#eq-actuarialFantasy}

The beta weights in the actuarial model reflect the relative weight to assign each predictor.
For instance, in predicting rushing yards, a player's historical performance is likely the strongest predictor, whereas injury history might be a relatively weaker predictor.
Thus, we might give historical performance a beta of 3 and injury history a beta of 1 to give a player's historical performance three times more weight than the player's injury history in predicting their rushing yards.
For generating the actuarial model, you could obtain the beta weights for each predictor from [multiple regression](#sec-multipleRegression), from [machine learning](#sec-machineLearning), or from prior research on the relative importance of each predictor.

@@

The challenge is what to do when the human and the algorithm disagree.
Hypothetically, humans reviewing and adjusting the results from the statistical algorithm could lead to more accurate prediction.
However, human input also could lead to the possibility or exacerbation of biased predictions.
In general, with very few exceptions, actuarial approaches are as accurate or more accurate than "expert" judgment [@Dawes1989; @Aegisdottir2006; @Baird2000; @Grove1996; @Grove2000].
This is also likely true with respect to predicting player performance in sports [@DenHartigh2018].
Moreover, the superiority of actuarial approaches to human judgment tends to hold even when the expert is given more information than the actuarial approach [@Dawes1989].
Allowing experts to override actuarial predictions consistently leads to lower predictive accuracy [@Garb2019].

There is sometimes a misconception that formulas cannot account for qualitative information.
However, that is not true.
Qualitative information can be scored or coded to be quantified so that it can be included in statistical formulas.
For instance, if an expert scout is able to meaningfully assess a player's [cognitive and motivational factors](#sec-evalCogMotivational) (i.e., the "[X factor](#sec-evalCogMotivational)" or "[intangibles](#sec-evalCogMotivational)"), the scout can score this across multiple players and include these data in the actuarial prediction formula.
For instance, the scout could use a rating scale (e.g., 1 = "poor"; 2 = "fair"; 3 = "good"; 4 = "very good"; 5 = "excellent") to code (i.e., translate) their qualitative judgment into a quantifiable rating that can be integrated with other information in the actuarial formula.
That said, the quality of predictions rests on the quality and relevance of the assessment information for the particular prediction decision.
If the assessment data are lousy, it is unlikely that a statistical algorithm (or a human for that matter) will make an accurate prediction:
"Garbage in, garbage out".
A statistical formula cannot rescue inaccurate assessment data.
